---
title: "An introduction to functional programming with R"
author: "Hon Hwang"
date: "2019 April"
output:
  beamer_presentation:
    colortheme: dove
    theme: Antibes
  includes: null
  slidy_presentation: default
  pdf_document:
    keep_tex: true
---

## A simple function
Write a function that adds 100 to its argument \texttt{x}


```{r}
add100 <- function(x) { return(x + 100) }
```

## Passing function to function

If a function can accept a number, a \texttt{data.frame} and other
data structures, can a function accept a function as well?

Write a function that

  - Accepts two arguments: $\texttt{z}$, $\texttt{fn}$,
    
  - Returns the result of calling `fn` with `z`. i.e., `fn(z)`

## Passing function to function

```{r}
customadd <- function(z, fn) {
  return(fn(z))
}

customadd(55, add100)
```

If _fn_ doesn't has a name, it's called an _anonymous function_.

In some other programming languages, _anonymous function_ are also called
_lambda_ function.


## Calling functions with `do.call` 

Using \texttt{do.call} to call an external function.

```{r}
customadd2 <- function(z, fn) {
  return( do.call(what = fn, args = list(x = z)))
}

customadd2(55, add100)
```

## Passing function to function

- First ability of functional programming language.

- Functions as \emph{first class} citizens.
  - Similar to other data types,
  - Functions can be supplied to functions.
  - Functions can return functions.

## Returning a function

Let's now get a function to return a function!
Write a function that

  - Accepts one argument: a value $\texttt{m}$,
    
  - Returns:
    - A function
      - Accepts one argument $\texttt{x}$
      - Returns $\texttt{m}$ plus $\texttt{x}$.

## Returning a function

```{r}
addfactory <- function(m) {
  return(
    function(x) { x + m }
  )
}
```

In _object-orientated programming_ world, these are called _factory_
objects.

## Returning a function

```{r}
add200 <- addfactory(200)
add200(10)

add500 <- addfactory(500)
add500(10)
```

## Presetting values of a function

A factory function that presets the number of data points to generate.

```{r}
rnorm_factory <- function(n) {
  return(
    function(mu, sd) {
      rnorm(n, mu, sd)
    }
  )
}

gen1k_rnorm <- rnorm_factory(1000)
mu2sd1 <- gen1k_rnorm(2, 1)
mu10sd1 <- gen1k_rnorm(10, 1)
```

## Presetting values of a function

```{r}
hist(c(mu2sd1, mu10sd1))
```
I am changing the \emph{behaviour} of a function.

## Presetting arguments using $\texttt{purrr}$

Using $\texttt{partial}$ function from $\texttt{purrr}$ package.
```{r, warning=FALSE, out.height="40%"}
library(purrr)

gen2k_rnorm <- partial(rnorm, n = 2000)
length(gen2k_rnorm())

hist(gen2k_rnorm())

```

##
```{r}
hist(gen2k_rnorm(mean = 50))
```

## The Map function

Applies a function to data

Rows oriented workflow slides 20 to 23

  - https://speakerdeck.com/jennybc/row-oriented-workflows-in-r-with-the-tidyverse?slide=20

## Walkthrough of \texttt{rnorm} examples from reference page

Acts like $\texttt{lapply}$ but can return results into different formats.

  - https://purrr.tidyverse.org/reference/map.html

## Eliminating loops

  - https://r4ds.had.co.nz/iteration.html#for-loops-vs.functionals
  
  - First `mtcars` example from
    https://purrr.tidyverse.org/reference/map.html
  
## Changing behaviours of functions

  - https://r4ds.had.co.nz/iteration.html#dealing-with-failure

## Mapping over rows

Create a `tibble` of params
```{r, warning=FALSE}
library(tibble)

params_tbl <- tibble(mu = c(1,5,10), sd = c(1,0.5,0.3))
purrr::pmap(params_tbl, function(mu,sd) {
  rnorm(10, mu, sd)
})
```

## For more in depth

R for data science

  - https://r4ds.had.co.nz/
    
`purrr` reference

  - https://purrr.tidyverse.org/index.html
    
`purrr` cheatsheat

  - https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf
    
Advanced R (2nd edition)

  - http://adv-r.had.co.nz/
    

## Functional programming continued - Closure

Scoping

```{r, scoping-intro}
x <- 10

addOneToNum <- function(num) {
  x <- num
  return(x + 1)
}

addOneToNum(10)

```

  - The first `x` resides in the global _environment_
  - The second `x` resides inside in the _environment_ of `addOneToNum` function 

## Closure - enclosing the _parent_ environment

From http://adv-r.had.co.nz/Functional-programming.html#closures

Create a factory function
```{r, scoping-advr-ex}
power <- function(exponent) {
  function(x) {
    x ^ exponent
  }
}
```

```{r, scoping-advr-ex-running}
square <- power(2)
square(2)

cube <- power(3)
cube(2)
```

## Closure - enclosing the _parent_ environment

  - Let's look at the _environment_ where `x` resides.
  - Add one extra line to factory function

```{r, scoping-advr-ex-getenv, warning=FALSE}
power <- function(exponent) {
  print(environment())
  function(x) {
    x ^ exponent
  }
}
```

## Closure - enclosing the _parent_ environment

> The parent environment of a closure is the execution environment of the
> function that created it 

```{r}
square2 <- power(2)

environment(square2)
```

## Closure - enclosing the _parent_ environment

Let's look at the value of `x` in each of the functions

```{r, scoping-advr-ex-loadpryr, warning=FALSE}
library(pryr)
```

##

```{r}
unenclose(square)
unenclose(cube)
```

## How to modify that `x` inside the closure

  - DANGER. If you need to modify or keep _stats_, use object-orientated
    functionalities of R.
  - Double arrow assignment (`<<-`)  

```{r, modstate-advr-ex}
new_counter <- function() {
  i <- 0
  function() {
    i <<- i + 1
    i
  }
}
```

##

```{r}
counter_one <- new_counter()
counter_two <- new_counter()

counter_one()
counter_two()

counter_one()
counter_one()
```

## MLE estimation using closures

http://adv-r.had.co.nz/Functionals.html#functionals-math

```{r, advr-ex-mle}
poisson_nll <- function(x) {
  n <- length(x)
  sum_x <- sum(x)
  function(lambda) {
    n * lambda - sum_x * log(lambda) # + terms not involving lambda
  }
}

poivals <- rpois(100, lambda = 32)
nll1 <- poisson_nll(poivals)

optimise(nll1, c(0, 100))$minimum
```

## Non-standard evaluation

http://adv-r.had.co.nz/Computing-on-the-language.html

```{r, out.height="40%"}
x <- seq(0, 2 * pi, length = 100)
sinx <- sin(x)
plot(x, sinx, type = "l")
```

## Expressions and values

  - Values: `x` and `sinx`
  - Expression: _code used to compute_ the value. E.g., `seq` and `sin` functions
  - To capture expressions, use `substitute` function.
  - `deparse` function turns expression captured by `substitute` into strings.
  
##

```{r, capture-exp}
f <- function(x) {
  return(substitute(x))
}

f(1:10)
deparse(f(1:10))
```

## `library` function

  - To avoid quotes when calling `library` function
```{r, eval=FALSE}
library(
  deparse(substitute(ggplot2)),
  character.only=TRUE
)
```

## `plot` function

From help of `deparse`

```{r, eval=FALSE}
myplot <-
     function(x, y) {
         plot(x, y, xlab = deparse(substitute(x)),
             ylab = deparse(substitute(y)))
     }
```

## Useful but complicated example using `ggplot`

```{r, ggplot-demo, out.height="50%"}
library(ggplot2)

ggplot(mtcars, aes(x=mpg, y=gear)) + geom_point()
```
Note `mpg` and `gear` are _not_ quoted; they are not characters

##

What if we have variables of the axis data?
```{r, ggplot-demo-axisvars, out.height="50%"}
xaxis <- "disp"
yaxis <- "drat"

ggplot(mtcars, aes(x=xaxis, y=yaxis)) + geom_point()
```

##

Turn variables into _symbols_ first then _unquote_

```{r, ggplot-demo-axisvars-symb, out.height="50%"}
xaxis_sym <- sym("disp")
yaxis_sym <- sym("drat")

ggplot(mtcars) + geom_point(aes(!!xaxis_sym, !!yaxis_sym))
```

## Simpler version (but deprecated)

See help on `aes_string`
https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf

```{r, ggplot-demo-axisvars-aesstr, out.height="50%"}
xaxis <- "disp"
yaxis <- "drat"

ggplot(mtcars) + geom_point(aes_string(x = xaxis, y = yaxis))
```

## Tidy evaluation in 5 minutes

https://youtu.be/nERXS3ssntw

## Domain specific language

Walk through parts of LaTeX and HTML code generation from

http://adv-r.had.co.nz/dsl.html#html

Hints on possibly how `RShiny` generates all those user interface.

## Domain specific language done by other functional programming language

Racket was created to make this web site on typography
https://practicaltypography.com/how-this-book-was-made.html

A great article on why use functional programming languages like Lisp
https://practicaltypography.com/why-racket-why-lisp.html

