---
title: "An introduction to functional programming with R"
author: "Hon Hwang"
date: "2019 April"
output:
  beamer_presentation:
    colortheme: dove
    theme: Antibes
  includes: null
  slidy_presentation: default
  pdf_document:
    keep_tex: true
---

## A simple function
Write a function that adds 100 to its argument \texttt{x}


```{r}
add100 <- function(x) { return(x + 100) }
```

## Passing function to function
Functions as _first class citizen_

If a function an accept a number, a \texttt{data.frame} and other
data structures, can a function accept a function as well?

Write a function that

  - Accepts two arguments: a value $\texttt{z}$ and a $\texttt{function}$,
    
  - Returns the result of calling the supplied function with the value of
    $\texttt{z}$.

## Passing function to function

```{r}
customadd <- function(z, fn) {
  return(fn(z))
}

customadd(10, add100)
```

## 

Using \texttt{do.call} to call an external function.

## Passing function to function
Write a function that calls the function in its argument, with the value
in its argument.

```{r}
customadd2 <- function(z, fn) {
  return( do.call(what = fn, args = list(x = z)))
}

customadd2(10, add100)
```

## Passing function to function

- First ability of functional programming language.

- Functions as \emph{first class} citizens.
  - Similar to other data types,
  - Functions can be supplied to functions.
  - Functions can return functions.

## Returning a function

Write a function that

  - Accepts one argument: a value $\texttt{m}$,
    
  - Returns:
    - A function that accepts one argument $\texttt{x}$
    - Returns $\texttt{m}$ plus $\texttt{x}$.

## Returning a function

```{r}
addfactory <- function(m) {
  return(
    function(x) { x + m }
  )
}

add200 <- addfactory(200)
add200(10)

add500 <- addfactory(500)
add500(10)
```

## Presetting values of a function

```{r}
rnorm_factory <- function(n) {
  return(
    function(mu, sd) { rnorm(n, mu, sd) }
  )
}

gen1k_rnorm <- rnorm_factory(1000)
mu2sd1 <- gen1k_rnorm(2, 1)
mu10sd1 <- gen1k_rnorm(10, 1)
```

## Presetting values of a function

```{r}
hist(c(mu2sd1, mu10sd1))
```
I am changing the \emph{behaviour} of a function.

## Presetting arguments using $\texttt{purrr}$

Using $\texttt{partial}$ function from $\texttt{purrr}$ package.
```{r, warning=FALSE, out.height="40%"}
library(purrr)

gen2k_rnorm <- partial(rnorm, n = 2000)
length(gen2k_rnorm())

hist(gen2k_rnorm())

```

##
```{r}
hist(gen2k_rnorm(mean = 50))
```

## The Map function

Applies a function to data

Rows oriented workflow slides 20 to 23

  - https://speakerdeck.com/jennybc/row-oriented-workflows-in-r-with-the-tidyverse?slide=20

## Walkthrough of \texttt{rnorm} examples from reference page

Acts like $\texttt{lapply}$ but can return results into different formats.

  - https://purrr.tidyverse.org/reference/map.html

## Eliminating loops

  - https://r4ds.had.co.nz/iteration.html#for-loops-vs.functionals
  
  - First `mtcars` example from
    https://purrr.tidyverse.org/reference/map.html
  
## Changing behaviours of functions

  - https://r4ds.had.co.nz/iteration.html#dealing-with-failure

## Mapping over rows

Create a `tibble` of params
```{r, warning=FALSE}
library(tibble)

params_tbl <- tibble(mu = c(1,5,10), sd = c(1,0.5,0.3))
purrr::pmap(params_tbl, function(mu,sd) {
  rnorm(10, mu, sd)
})
```

## For more in depth

R for data science

  - https://r4ds.had.co.nz/
    
`purrr` reference

  - https://purrr.tidyverse.org/index.html
    
`purrr` cheatsheat

  - https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf
    
Advanced R (2nd edition)

  - http://adv-r.had.co.nz/
    
